# R Programming Basics

Before we get started, you will need to know the basics of matrix manipulation in the R programming language:

- Generally matrices are entered in as one vector, which R then breaks apart into rows and columns in they way that you specify (with nrow/ncol). The default way that R reads a vector into a matrix is down the columns. To read the data in across the rows, use the byrow=TRUE option). This is only relevant if you're entering matrices from scratch.

```{r}
Y=matrix(c(1,2,3,4),nrow=2,ncol=2)
Y
X=matrix(c(1,2,3,4),nrow=2,ncol=2,byrow=TRUE)
X
```

- The standard multiplication symbol, '\*', will unfortunately provide unexpected results if you are looking for matrix multiplication. '\*' will multiply matrices _elementwise_.  In order to do matrix multiplication, the function is '%\*%'.

```{r}
X*X
X%*%X
```

- To transpose a matrix or a vector $\X$, use the function t($\X$).
```{r}
t(X)
```

- R indexes vectors and matrices starting with $i=1$ (as opposed to $i=0$ in python).
- X[i,j] gives element $\X_{ij}$. You can alter individual elements this way.

```{r}
X[2,1]
X[2,1]=100
X
```

- To create a vector of all ones, $\e$, use the ```  rep()``` function

```{r}
e=rep(1,5)
e
```

- To compute the mean of a vector, use the mean function. To compute the column means of a matrix (or data frame), use the ```  colMeans() ```  function. You can also use the ``` apply``` function, which is necessary if you want column standard deviations (```  sd() ``` function). ```  apply(X,dim,function)``` applies the specified function to the specified dimension ```  dim``` (1 for rows, 2 for columns) of the matrix or data frame X.
```{r}
# Start by generating random ~N(0,1) data:
A=replicate(2,rnorm(5))
colMeans(A)
# (Why aren't the means close to zero?)
A=replicate(2,rnorm(100))
colMeans(A)
#LawOfLargeNumbers.

apply(A,2,sd)
# To apply a "homemade function" you must create it as a function
# Here we apply a sum of squares function for the first 5 rows of A:
apply(A[1:5, ],1,function(x) x%*%x)
# Here we center the data by subtracting the mean vector:
B=apply(A,2,function(x) x-mean(x))
colMeans(B)
# R doesn't tell you when things are zero to machine precision. "Machine zero" in
# R is given by the internal variable .Machine$double.eps
colMeans(B) < .Machine$double.eps
```

- To invert a matrix, use the ```  solve()``` command.

```{r}
Xinv=solve(X)
X%*%Xinv
```

- To determine size of a matrix, use the ```  dim()``` function. The result is a vector with two values: ```  dim(x)[1]``` provides the number of rows and ```  dim(x)[2]``` provides the number of columns. You can label rows/columns of a matrix using the ```  rownames()``` or ```  colnames()``` functions.

```{r}
dim(A)
nrows=dim(A)[1]
ncols=dim(A)[2]
colnames(A)=c("This","That")
A[1:5, ]
```

- Most arithmetic functions you apply to a vector act elementwise. In R, $\x^2$ will be a vector containing the square of the elements in $\x$. You can add a column to a matrix (or a data frame) by using the ```  cbind()``` function.

```{r}
# Add a column containing the square of the second column
A=cbind(A,A[ ,2]^2)
colnames(A)
colnames(A)[3]="That Squared"
colnames(A)
```

- You can compute vector norms using the ```  norm() ``` function. Unfortunately, the default norm is _not_ the $2$-norm (it should be!) so we must specify the ``` type="2" ``` as the second argument to the function.

```{r}
x=c(1,1,1)
y=c(1,0,0)
norm(x,type="2")
# It's actually fewer characters to work from the equivalent definition:
sqrt(x%*%x)
norm(y,type="2")
norm(x-y,type="2")
```

You'll learn many additional R techniques throughout this course, but our strategy in this text will be to pick them up as we go as opposed to trying to remember them from the beginning.

